---
title: "Random Forests"
format: 
  revealjs:
    theme: simple
    transition: slide
    slide-number: true
    chalkboard: true
execute:
  echo: false
  warning: false
  message: false
---


## Building The Model {.smaller}

:::: {.columns}

```{r}
library(randomForest)
library(caret)
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(reprtree)

data(iris)
```

::: {.column width="60%"}
```{r}
#| label: Classification Initial Model
set.seed(42)

trainIndex = sample(1:nrow(iris), 0.8 * nrow(iris))

trainData = iris[trainIndex, ]
testData = iris[-trainIndex, ]

rf_model = randomForest(Species ~ ., data = trainData, type="classification", importance=T, proximity=T)

print(rf_model)
```
:::
::: {.column width="40%"}
Out-of-Bag Error Rate: $\frac{1}{n} \sum_{i=1}^{n} \mathbf{1}\left( y_i \neq \hat{y}_i^{\text{OOB}} \right)$

- If the majority vote prediction for observation $i$ using only the trees that did not feature observation $i$ in their training subset is not equal to the true label for observation $i$, then we set to 1. 0 otherwise.
- Hyperparameters: Number of trees and No. of variables tried at each split (mtry)

:::
::::

## Singular Decision Tree

```{r}
plot.getTree(rf_model, k = 1)
```

## Feature Influence{.smaller}


```{r fig.width=9, fig.height=5}
imp = importance(rf_model)

imp_df = as.data.frame(imp) |>
  tibble::rownames_to_column("Feature") |>
  pivot_longer(
    cols = c(MeanDecreaseAccuracy, MeanDecreaseGini),
    names_to  = "Metric",
    values_to = "Importance"
  ) |>
  mutate(
    Metric  = recode(Metric,
      MeanDecreaseAccuracy = "Mean Decrease Accuracy",
      MeanDecreaseGini = "Mean Decrease Gini"
    ),
    Feature = fct_reorder(Feature, Importance, .fun = mean)
  )

ggplot(imp_df, aes(x = Importance, y = Feature, fill = Metric)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.55, alpha = 0.88) +
  facet_wrap(~Metric, scales = "free_x") +
  scale_fill_manual(values = c("#2C7BB6", "#D7191C")) +
  labs(
    title = "Feature Importance",
    subtitle = "Accuracy: permutation-based  |  Gini: impurity-based",
    x = "Importance Score",
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Learning Curve — OOB Error vs. Trees {.smaller}

```{r fig.width=9, fig.height=5}
plot(rf_model, type = "l", main = "OOB Error vs. Number of Trees")
```


## Model Evaluation {.smaller}

```{r}
predictions <- predict(rf_model, testData)
confusionMatrix(predictions, testData$Species)
```

# Hyperparameter Tuning

## Random Search for Hyperparameter Tuning

```{r}
ctrl_random = trainControl(method = "cv", number = 5, search = "random")

set.seed(123)
random_model = train(
  Species ~ .,
  data = trainData,
  method = "ranger",
  tuneLength = 20,   
  trControl = ctrl_random
)

ntree_vals = c(50, 100, 200, 300, 500)
oob_errors = sapply(ntree_vals, function(n) {
  rf <- randomForest(Species ~ ., data = trainData, ntree = n, mtry = 3, nodesize = 14)
  rf$err.rate[n, "OOB"]
})

cat("\nBest Hyperparameters (Random Search):\n")
print(random_model$bestTune)

best_ntree = ntree_vals[which.min(oob_errors)]
cat("Best ntree:", best_ntree, "\n")

```

## Did we improve?

```{r}
set.seed(42)
rf_model_tune <- randomForest(Species ~ ., data = trainData, type="classification", importance=T, proximity=T, ntree=50, mktry=3, nodesize=14)

print(rf_model_tune)
```

