---
title: "Random Forests"
author: "Ryan McMurray, Melanie Jensen, Lucca Coelho"
format: 
  revealjs:
    theme: simple
    transition: slide
    slide-number: true
    chalkboard: true
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

# Mathematical Foundation

## Introduction

Random forests are an ensemble learning method 

- They combine many decision trees into a single model  
- Used for regression and classification
- Used with complex, nonlinear data

## Key Idea

- Build many different trees  
- Make them decorrelated
- Average their predictions

### Why It Works

- Individual trees → high variance  
- Averaging → reduces variance
- Result → more stable and accurate model

## Regression Trees
For regression, a tree partitions the feature space into regions 
$R_1, R_2, \dots, R_J$
, and predicts:

$\hat{f}(x) = \sum_{j=1}^{J} c_j \mathbf{1}(x \in R_j)$

where 
$c_j$ 
is typically the mean of the response values in region 
$R_j$
.

Splits are chosen to minimize residual sum of squares (RSS):

$\sum_{i: x_i \in R_1} (y_i - \bar{y}_{R_1})^2 + \sum_{i: x_i \in R_2} (y_i - \bar{y}_{R_2})^2$

## Classification Trees
For classification, trees predict class labels using majority vote within each region.

Splits are chosen to reduce impurity, commonly using:

- **Gini Index**:

$G = \sum_{k=1}^{K} p_k (1 - p_k)$

- **Entropy**:

$H = -\sum_{k=1}^{K} p_k \log p_k$

where 
$p_k$
is the proportion of class
$k$
in a node.

## Random Forest Algorithm

A random forest builds 
$B$
trees using two sources of randomness:

### 1. Bootstrap Sampling
For each tree 
$b = 1, \dots, B$
, draw a bootstrap sample from the data.

### 2. Random Feature Selection
At each split, select a random subset of $m$ features (with $m < p$) and choose the best split only among those features.

## Why Random Forests Work

Let each tree have variance 
$\sigma^2$ 
and pairwise correlation 
$\rho$
. Then the variance of the ensemble is approximately:

$\text{Var}(\hat{f}) \approx \rho \sigma^2 + \frac{1 - \rho}{B} \sigma^2$

Key insight:

- Increasing  
$B$
reduces variance
- Reducing correlation 
$\rho$ 
(via feature randomness) is crucial

## Assumptions

Random forests make fewer assumptions than parametric models, but still rely on:

- Independent Observation
- The training set is representative of the population
- Variables must provide information on the dependent variable
- Predictions must be within the range of the training data

## Out-of-Bag (OOB) Error Estimation

Each tree is trained on a bootstrap sample, leaving about 
$\approx 1/3$ 
of observations unused.

For each observation, predictions from trees where it was not included are averaged:

$\hat{f}_{\text{OOB}}(x_i) = \frac{1}{|\mathcal{B}_i|} \sum_{b \notin \mathcal{B}_i} T_b(x_i)$

This provides an unbiased estimate of generalization error.

## Shortcomings

- Lack of Interpretability
- Computational Cost
- Bias in High-Dimensional Sparse Settings
- Poor Extrapolation
- Correlated Features

## Real-World Applications

- **Healthcare**
  - Predicting disease risk
  - Identifying important risk factors  

- **Finance**
  - Credit scoring (loan approval decisions)  
  - Fraud detection in transactions  

- **Marketing**
  - Customer churn prediction  
  - Targeted advertising  

## Summary

Random forests improve upon decision trees by:

- Reducing variance through averaging
- Reducing correlation via random feature selection

They are a flexible, powerful tool that performs well in practice with minimal tuning, making them a strong choice for both regression and classification tasks.


# Regression 

## AMES Housing Data
- 2,930 homes, 81 variables
- Predicting Sale_Price from everything else

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(randomForest)
library(AmesHousing)
library(GGally)
library(kableExtra)

ames <- make_ames()

head(ames) |>
    select(Sale_Price, Gr_Liv_Area, Overall_Qual, Year_Built, 
           Neighborhood, Total_Bsmt_SF) |>
    kable() |>
    kable_styling(font_size = 20)

set.seed(2026)
split <- sample(nrow(ames), size = 0.8 * nrow(ames))
train <- ames[split, ]
test <- ames[-split, ]

```

## Correlation Plot
```{r}
#| label: pairplot

num_ames <- ames |> select(where(is.numeric))
cors <- cor(num_ames, use = "complete.obs")["Sale_Price", ]
top_vars <- names(sort(abs(cors), decreasing = TRUE)[1:8])

ggpairs(
  num_ames[, top_vars],
  upper = list(continuous = wrap("cor", size = 3)),
  lower = list(continuous = wrap("points", alpha = 0.2, size = 0.5)),
  diag  = list(continuous = wrap("densityDiag")),
  title = "Pairplot: Top 8 Numeric Variables Correlated with Sale Price"
) +
  theme_minimal(base_size = 10) +
  theme(strip.text.y = element_text(angle = 0))

```


## Random Forest Model 
```{r}
#| label: random-forest

set.seed(2026)

rf_model <- randomForest(Sale_Price ~ ., data = train, ntree = 500, mtry = floor((ncol(train) - 1) / 3), importance = TRUE)

rf_model

```

## Random Forest Tree Tuning

```{r}
#| label: rf-tree-tuning

tibble(
  trees = 1:rf_model$ntree,
  oob_mse = rf_model$mse
) |>
  ggplot(aes(x = trees, y = oob_mse)) +
  geom_line(color = "black", linewidth = 0.8) +
  labs(
    x = "Number of trees",
    y = "OOB MSE"
  ) +
  theme_minimal()
```

## Random Forest mtry Tuning

```{r}
#| label: rf-mtry-tuning-code
#| eval: false

set.seed(2026)
tuneRF(
    x = train |> select(-Sale_Price),
    y = train$Sale_Price,
    ntreeTry = 200,
    stepFactor = 1.5,
    improve = 0.01,
    trace = FALSE,
    plot = TRUE
)

```

```{r}
#| label: rf-mtry-tuning-output
#| echo: false

set.seed(2026)
tuneRF(
    x = train |> select(-Sale_Price),
    y = train$Sale_Price,
    ntreeTry = 200,
    stepFactor = 1.5,
    improve = 0.01,
    trace = FALSE,
    plot = TRUE
)

```

## Tuned Random Forest Model


```{r}
#| label: tuned-random-forest

set.seed(2026)
tuned_rf_model <- randomForest(Sale_Price ~ ., data = train, ntree = 200, mtry = 26, importance = TRUE)

tuned_rf_model

```

## Model Evaluation

```{r}
#| label: rf-evaluation-setup

test_preds <- predict(tuned_rf_model, newdata = test)
test_actual <- test$Sale_Price

test_rmse <- sqrt(mean((test_actual - test_preds)^2))
test_r2 <- 1 - sum((test_actual - test_preds)^2) / 
               sum((test_actual - mean(test_actual))^2)
oob_rmse <- sqrt(tuned_rf_model$mse[tuned_rf_model$ntree])

```

::::{.columns}
::: {.column width="20%"}

```{r}
#| label: rf-evaluation-table

tibble(
  Metric = c("Test RMSE", "Test R²", "OOB RMSE"),
  Value = c(
    paste0("$", format(round(test_rmse), big.mark = ",")),
    round(test_r2, 3),
    paste0("$", format(round(oob_rmse), big.mark = ","))
  )
) |>
  kable() |>
  kable_styling(font_size = 20, full_width = FALSE)

```

:::
::: {.column width="40%"}

```{r}
#| label: rf-evaluation-plot
#| fig-width: 4
#| fig-height: 3

tibble(actual = test_actual, predicted = test_preds) |>
  ggplot(aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, color = "black") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Actual Sale Price",
    y = "Predicted Sale Price"
  ) +
  theme_minimal()

```

:::
::: {.column width="40%"}

```{r}
#| label: rf-residual_plot
#| fig-width: 4
#| fig-height: 3

tibble(
  predicted = test_preds,
  residual = test_actual - test_preds
) |>
  ggplot(aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.4, color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(x = "Predicted Sale Price", y = "Residual (Actual − Predicted)") +
  theme_minimal()

```

:::
::::

## Feature Importance

:::: {.columns}
::: {.column width="60%"}

```{r}
#| label: feature-importance
#| fig-width: 14
#| fig-height: 14

importance_df <- importance(tuned_rf_model) |>
  as.data.frame() |>
  tibble::rownames_to_column("variable") |>
  arrange(desc(`%IncMSE`))

ggplot(importance_df, aes(x = `%IncMSE`, y = reorder(variable, `%IncMSE`))) +
  geom_col(fill = "steelblue") +
  labs(
    x = "% Increase in MSE when variable is permuted",
    y = NULL,
    title = "Variable importance (permutation method)"
  ) +
  theme_minimal()

```

:::
::: {.column width="40%"}

<div style="font-size: 0.6em;">

When we scramble *variable* values in the OOB data, the model's MSE increases by *value*%.

Variables at the bottom either carry redundant information already captured by other predictors, or have no real relationship to price at all. Either way, the model isn't using them meaningfully. When we permute these unused variables, the MSE barely changes — it wobbles randomly above or below the original value. Negative values aren't evidence that the variable was harmful; they're just what noise looks like around zero when a variable has no true importance.

</div>

:::
::::


# Classification

## Building The Model {.smaller}

:::: {.columns}

```{r}
library(tidyverse)
library(randomForest)
library(AmesHousing)
library(GGally)
library(kableExtra)
library(caret)
library(rpart)
library(rpart.plot)

data(iris)
```

::: {.column width="60%"}
```{r}
#| label: Classification Initial Model
set.seed(42)

trainIndex = sample(1:nrow(iris), 0.8 * nrow(iris))

trainData = iris[trainIndex, ]
testData = iris[-trainIndex, ]

rf_model = randomForest(Species ~ ., data = trainData, type="classification", importance=T, proximity=T)

print(rf_model)
```
:::
::: {.column width="40%"}
Out-of-Bag Error Rate: $\frac{1}{n} \sum_{i=1}^{n} \mathbf{1}\left( y_i \neq \hat{y}_i^{\text{OOB}} \right)$

- If the majority vote prediction for observation $i$ using only the trees that did not feature observation $i$ in their training subset is not equal to the true label for observation $i$, then we set to 1. 0 otherwise.
- Hyperparameters: Number of trees and No. of variables tried at each split (mtry)

:::
::::

## Singular Decision Tree

```{r}
#| label: single-tree
set.seed(2026)

tree_model <- rpart(
  Species ~ .,
  data = trainData,
  method = "anova")


rpart.plot(tree_model, main = "Single Decision Tree Approximation")
```

## Feature Influence{.smaller}


```{r fig.width=9, fig.height=5}
imp = importance(rf_model)

imp_df = as.data.frame(imp) |>
  tibble::rownames_to_column("Feature") |>
  pivot_longer(
    cols = c(MeanDecreaseAccuracy, MeanDecreaseGini),
    names_to  = "Metric",
    values_to = "Importance"
  ) |>
  mutate(
    Metric  = recode(Metric,
      MeanDecreaseAccuracy = "Mean Decrease Accuracy",
      MeanDecreaseGini = "Mean Decrease Gini"
    ),
    Feature = fct_reorder(Feature, Importance, .fun = mean)
  )

ggplot(imp_df, aes(x = Importance, y = Feature, fill = Metric)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.55, alpha = 0.88) +
  facet_wrap(~Metric, scales = "free_x") +
  scale_fill_manual(values = c("#2C7BB6", "#D7191C")) +
  labs(
    title = "Feature Importance",
    subtitle = "Accuracy: permutation-based  |  Gini: impurity-based",
    x = "Importance Score",
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Learning Curve — OOB Error vs. Trees {.smaller}

```{r fig.width=9, fig.height=5}
plot(rf_model, type = "l", main = "OOB Error vs. Number of Trees")
```


## Model Evaluation {.smaller}

```{r}
predictions <- predict(rf_model, testData)
confusionMatrix(predictions, testData$Species)
```

# Hyperparameter Tuning

## Random Search for Hyperparameter Tuning

```{r}
ctrl_random = trainControl(method = "cv", number = 5, search = "random")

set.seed(123)
random_model = train(
  Species ~ .,
  data = trainData,
  method = "ranger",
  tuneLength = 20,   
  trControl = ctrl_random
)

ntree_vals = c(50, 100, 200, 300, 500)
oob_errors = sapply(ntree_vals, function(n) {
  rf <- randomForest(Species ~ ., data = trainData, ntree = n, mtry = 3, nodesize = 14)
  rf$err.rate[n, "OOB"]
})

cat("\nBest Hyperparameters (Random Search):\n")
print(random_model$bestTune)

best_ntree = ntree_vals[which.min(oob_errors)]
cat("Best ntree:", best_ntree, "\n")

```

## Did we improve?

```{r}
set.seed(42)
rf_model_tune <- randomForest(Species ~ ., data = trainData, type="classification", importance=T, proximity=T, ntree=50, mtry=2, nodesize=14)

print(rf_model_tune)
```



# Questions