---
title: "Random Forests"
format: 
  revealjs:
    theme: simple
    transition: slide
    slide-number: true
    chalkboard: true
---

# Random Forest Regression

## AMES Housing Data
- 2,930 homes, 81 variables
- Predicting Sale_Price from everything else

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(randomForest)
library(AmesHousing)
library(GGally)
library(kableExtra)

ames <- make_ames()

head(ames) |>
    select(Sale_Price, Gr_Liv_Area, Overall_Qual, Year_Built, 
           Neighborhood, Total_Bsmt_SF) |>
    kable() |>
    kable_styling(font_size = 20)

set.seed(2026)
split <- sample(nrow(ames), size = 0.8 * nrow(ames))
train <- ames[split, ]
test <- ames[-split, ]

```

## Correlation Plot
```{r}
#| label: pairplot

num_ames <- ames |> select(where(is.numeric))
cors <- cor(num_ames, use = "complete.obs")["Sale_Price", ]
top_vars <- names(sort(abs(cors), decreasing = TRUE)[1:8])

ggpairs(
  num_ames[, top_vars],
  upper = list(continuous = wrap("cor", size = 3)),
  lower = list(continuous = wrap("points", alpha = 0.2, size = 0.5)),
  diag  = list(continuous = wrap("densityDiag")),
  title = "Pairplot: Top 8 Numeric Variables Correlated with Sale Price"
) +
  theme_minimal(base_size = 10) +
  theme(strip.text.y = element_text(angle = 0))

```

## Linear vs. random forest 

<div style="font-size: 0.6em;">

::::{.columns}
::: {.column width="50%"}

### Linear regression 

**Strengths:**

- Interpretable coefficients
- Inference tools (p-values, CIs)
- Extrapolates beyond training data
- Fast to fit and predict

**Weaknesses:**

- Strict distributional assumptions
- Can't capture non-linearity without feature engineering
- Multicollinearity destabilizes coefficients
- Sensitive to high-cardinality categoricals

:::
::: {.column width="50%"}

### Random forest

**Strengths:**

- No distributional assumptions
- Captures non-linearity and interactions automatically
- Handles multicollinearity gracefully
- Built-in feature importance and OOB validation

**Weaknesses:**

- Less interpretable (no coefficients)
- Cannot extrapolate beyond training range
- Computationally heavier
- Harder to perform formal inference

:::
::::

</div>

## Random Forest Model 
```{r}
#| label: random-forest

set.seed(2026)

rf_model <- randomForest(Sale_Price ~ ., data = train, ntree = 500, mtry = floor((ncol(train) - 1) / 3), importance = TRUE)

rf_model

```

## Random Forest Tree Tuning

```{r}
#| label: rf-tree-tuning

tibble(
  trees = 1:rf_model$ntree,
  oob_mse = rf_model$mse
) |>
  ggplot(aes(x = trees, y = oob_mse)) +
  geom_line(color = "black", linewidth = 0.8) +
  labs(
    x = "Number of trees",
    y = "OOB MSE"
  ) +
  theme_minimal()
```

## Random Forest mtry Tuning

```{r}
#| label: rf-mtry-tuning-code
#| eval: false

set.seed(2026)
tuneRF(
    x = train |> select(-Sale_Price),
    y = train$Sale_Price,
    ntreeTry = 200,
    stepFactor = 1.5,
    improve = 0.01,
    trace = FALSE,
    plot = TRUE
)

```

```{r}
#| label: rf-mtry-tuning-output
#| echo: false

set.seed(2026)
tuneRF(
    x = train |> select(-Sale_Price),
    y = train$Sale_Price,
    ntreeTry = 200,
    stepFactor = 1.5,
    improve = 0.01,
    trace = FALSE,
    plot = TRUE
)

```

## Tuned Random Forest Model


```{r}
#| label: tuned-random-forest

set.seed(2026)
tuned_rf_model <- randomForest(Sale_Price ~ ., data = train, ntree = 200, mtry = 26, importance = TRUE)

tuned_rf_model

```

## Model Evaluation

```{r}
#| label: rf-evaluation-setup

test_preds <- predict(tuned_rf_model, newdata = test)
test_actual <- test$Sale_Price

test_rmse <- sqrt(mean((test_actual - test_preds)^2))
test_r2 <- 1 - sum((test_actual - test_preds)^2) / 
               sum((test_actual - mean(test_actual))^2)
oob_rmse <- sqrt(tuned_rf_model$mse[tuned_rf_model$ntree])

```

::::{.columns}
::: {.column width="20%"}

```{r}
#| label: rf-evaluation-table

tibble(
  Metric = c("Test RMSE", "Test R²", "OOB RMSE"),
  Value = c(
    paste0("$", format(round(test_rmse), big.mark = ",")),
    round(test_r2, 3),
    paste0("$", format(round(oob_rmse), big.mark = ","))
  )
) |>
  kable() |>
  kable_styling(font_size = 20, full_width = FALSE)

```

:::
::: {.column width="40%"}

```{r}
#| label: rf-evaluation-plot
#| fig-width: 4
#| fig-height: 3

tibble(actual = test_actual, predicted = test_preds) |>
  ggplot(aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, color = "black") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Actual Sale Price",
    y = "Predicted Sale Price"
  ) +
  theme_minimal()

```

:::
::: {.column width="40%"}

```{r}
#| label: rf-residual_plot
#| fig-width: 4
#| fig-height: 3

tibble(
  predicted = test_preds,
  residual = test_actual - test_preds
) |>
  ggplot(aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.4, color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(x = "Predicted Sale Price", y = "Residual (Actual − Predicted)") +
  theme_minimal()

```

:::
::::

## Feature Importance

:::: {.columns}
::: {.column width="60%"}

```{r}
#| label: feature-importance
#| fig-width: 14
#| fig-height: 14

importance_df <- importance(tuned_rf_model) |>
  as.data.frame() |>
  tibble::rownames_to_column("variable") |>
  arrange(desc(`%IncMSE`))

ggplot(importance_df, aes(x = `%IncMSE`, y = reorder(variable, `%IncMSE`))) +
  geom_col(fill = "steelblue") +
  labs(
    x = "% Increase in MSE when variable is permuted",
    y = NULL,
    title = "Variable importance (permutation method)"
  ) +
  theme_minimal()

```

:::
::: {.column width="40%"}

<div style="font-size: 0.6em;">

When we scramble *variable* values in the OOB data, the model's MSE increases by *value*%.

Variables at the bottom either carry redundant information already captured by other predictors, or have no real relationship to price at all. Either way, the model isn't using them meaningfully. When we permute these unused variables, the MSE barely changes — it wobbles randomly above or below the original value. Negative values aren't evidence that the variable was harmful; they're just what noise looks like around zero when a variable has no true importance.

</div>

:::
::::
